Emulation of high multiplicity matrix elements for electron-positron and hadron-hadron colliders

### 2022-09-14 intro.tex
wrote: cross-section, matrix elements, divergences

\section{Introduction to quantum chromodynamics}
    \subsection{The QCD Lagrangian}
    Quantum chronodynamics (QCD) is the sector of the SM that
    describes the strong interaction. 
    QCD is a non-Abelian gauge theory with gauge group
    $\mathrm{SU}(3)_{c}$ where the charge is named colour.
    The gauge and fermion part of the QCD Lagrangian is
    \begin{equation}\label{eqn:L_QCD}
        \mathcal{L}_{\mathrm{QCD}} = -\dfrac{1}{4}F^{a}_{\mu\nu}F^{a, \, \mu\nu}
        + \sum_{f} \bar{\psi}_{i}^{f}(\mathrm{i}\slashed{D}_{ij} - \delta_{ij}m_{f})\psi_{j}^{f} \, ,
    \end{equation}
    where twice repeated indices are summed over.
    The fields $\psi_{i}^{f}$ ($\bar{\psi}_{i}^{f}$) are the fermions (antifermions),
    representing quarks (antiquarks) with flavours $f$: 
    up $u$, down $d$, strange $s$, charm $c$, top $t$, and bottom $b$,
    with masses $m_{f}$. They transform under the fundamental
    representation with indices $i, j \in \{1, 2, 3\}$, named colour indices.
    The gauge fields $A^{a}_{\mu}$, corresponding to gluons, 
    appear in the quark covariant derivative
    \begin{equation}\label{eqn:covariant_deriv}
        (D_{\mu})_{ij} = \delta_{ij}\partial_{\mu} - ig_{\mathrm{s}}T^{a}_{ij}A^{a}_{\mu} \, ,
    \end{equation}
    and the gauge field strength tensor
    \begin{equation}\label{eqn:F_munu}
        F^{a}_{\mu\nu} = \partial_{\mu}A^{a}_{\nu} - \partial_{\nu}A^{a}_{\mu} + g_{\mathrm{s}} f^{abc}A^{b}_{\mu}A^{c}_{\nu} \, .
    \end{equation}
    Gauge fields transform under the adjoint representation, 
    which is an $N_{c}^{2}-1$ dimensional representation, meaning
    the adjoint indices $a,b,c \in \{1,...,8\}$. $T^{a}_{ij}$ are the
    group generators in a fundamental representation. In SU(3) it is
    common to write the group generators as
    \begin{equation}\label{eqn:group_generators}
        T^{a}_{ij} = \dfrac{1}{2}\lambda^{a}_{ij}
    \end{equation}
    where $\lambda^{a}_{ij}$ are the Gell-Mann matrices \cite{Gell-Mann:1962yej}.
    The generators of the group have to obey the Lie algebra
    \begin{equation}\label{eqn:lie_algebra}
        [T^{a}, T^{b}] = \mathrm{i}f^{abc}T^{c} \, ,
    \end{equation}
    where $f^{abc}$ are the structure constants of SU(3).
    The gauge coupling of the group $g_{\mathrm{s}}$
    is a dimensionless free parameter of the theory.
    It is common to use the strong coupling constant
    instead, 
    \begin{equation}\label{eqn:alpha_s}
        \alpha_{\mathrm{s}} = \dfrac{g_{\mathrm{s}}^{2}}{4\pi} \, .
    \end{equation}
    The strong coupling constant, is in fact not constant, but
    is dependent on the energy scale of the process, $\mu$. This
    is due to the process of renormalisation which will be discussed
    in section ...

@@

    \subsection{QCD Feynman rules}
    Now that we have written down the QCD Lagrangian,
    we can examine the terms to write down how the
    gauge bosons interact with the fermions. A convenient
    way to do this is with Feynman rules, which can be
    pieced together to form Feynman diagrams. We will see
    that Feynman diagrams are a simple, but powerful, tool
    to systematically build all the possible ways a process
    can occur. This will be discussed further in the
    scattering amplitude section.

@@

    Note that we have only written down the vertices
    and propagators for the physical gluons and quarks,
    ghosts also have associated Feynman rules but we do
    not collect those here. A full list of Feynman rules
    in the SM can be found at Ref. \cite{Romao:2012pq}.

    Now that we have the Feynman rules for the theory
    we can move on to discuss scattering amplitudes
    and cross-sections.

\section{Factorisation theorem}
    At the LHC, collisions occur between two protons,
    which have constituent quarks and gluons, collectively
    named partons.
    The application of QCD to describe phenomenon
    in these proton-proton collisions rests on the use of
    the factorisation theorem which enables the
    separation of soft and hard energy scales.
    Within the proton, there are composite quarks
    and gluons that are constantly absorbed and
    emitted on a timescale inversely proportional
    to the mass of the proton (or hadron in general).
    This fluctuation timescale is much longer
    than the timescale of the interaction between
    an incoming probing parton from another proton in a
    highly energetic collision, like at the LHC.
    During the collision, or the hard scattering,
    the probe is able to interact with a
    parton that is effectively frozen.
    The probing parton knows nothing of the
    proton being probed, except for the fact
    that it collided with a parton carrying
    a fraction of the proton momentum. The
    distribution of partons within a parton
    is process independent and a fundamental
    property of the proton, meaning it can be
    separated from the actual scattering between
    the partons.

    This heuristic argument motivates the form of
    the factorisation equation, where the production of
    particles from a hadronic collision, the hadronic
    cross-section, can be
    written as
    \begin{equation}\label{eqn:hadronic_cs}
        \sigma_{AB \rightarrow n} = \sum_{a, b}\int_{0}^{1} \mathrm{d}x_{a}\mathrm{d}x_{b} \; f_{a/A}(x_{a},\mu_{\mathrm{F}})f_{b/B}(x_{b}, \mu_{\mathrm{F}}) \; \hat{\sigma}_{ab \rightarrow n}(Q, \mu_{\mathrm{F}}, \mu_{\mathrm{R}}) + \mathcal{O}\left(\dfrac{\Lambda_{\mathrm{QCD}}}{Q}\right)\, ,
    \end{equation}
    where $a$ is the parton identified from hadron $A$,
    and $b$ is the parton identified from hadron $B$.
    This factorisation is not exact as indicated by the
    correction term inversely proportional to characteristic
    hard scale $Q$. However, for high energy collisions, where
    $Q \gg \Lambda_{\mathrm{QCD}}$, this term and higher order
    terms are suppresed.
    In fact, the factorisation equation has only been
    proven for a few specific cases (DIS etc, cite here).

    $f_{a/h}(x, \mu)$ are parton distribution functions
    (PDFs) which depend on the momenta fractions $x$ parton $a$
    has with respect to its parent hadron $h$, at a scale $\mu$,
    usually taken to be the factorisation scale $\mu_{\mathrm{F}}$,
    the interface between soft and hard physics.
    The interpretation of PDFs at leading order are
    as probability distributions. $f_{a/h}(x, \mu)$ is
    the probability of finding parton $a$ within $h$
    carrying a momentum fraction $x$ at the energy scale $\mu$.
    PDFs are non-perturbative objects that encapsulate
    the soft effects in a scattering process below energy
    $\mu_{\mathrm{F}}$. They are non-perturbative because
    they are determined by fitting to experimental data,
    instead of calculated in perturbation theory.
    Multiple groups release PDF fits (cite groups here).
    In practice, these PDF fits
    are accessed through the LHAPDF interface (cite here).
    The evolution of PDFs between factorisation scales is possible
    through the use of the Dokshitzer-Gribov-Lipatov-Altarelli-Parisi (DGLAP)
     equations (cite here).

    $\hat{\sigma}_{ab \rightarrow n}(Q, \mu_{\mathrm{F}}, \mu_{\mathrm{R}})$
    is the partonic cross-section which descibes the interaction of
    partons $a$ and $b$ scattering into $n$ particles,
    with the scattering occurs at energy scale $Q$.
    Notice that the partonic cross-section depends on
    both the factorisation scale, $\mu_{\mathrm{F}}$ and
    the renormalisation scale, $\mu_{\mathrm{R}}$.
    {\color{red} Need to mention how these scales
    are unphysical but there is a residual dependence
    on them at fixed-order calculations. Scale
    variations are a conventional way to determine
    the effects of higher order terms.
    Renormalisation will be discussed in Chapter 2
    when divergences are introduced.}

    The discussion of how $\hat{\sigma}_{ab \rightarrow n}$
    is computed is discussed in detail in the next section.

\section{Scattering amplitudes and cross-sections}
    {\color{red} Cross-sections here are all
    partonic cross-sections. Need to talk about
    how they can be calculated order by order
    in perturbation theory with Feynman diagrams.}

    In collider experiments we typically collide
    two beams, consisting of bundles of energetic
    particles, and analyse the resulting products
    of these collisions. To model collisions in
    a collider experiment, we look at the specific
    case of two particles colliding into $n$ particles.
    To mathematically describe these
    scattering processes, scattering amplitudes are used.
    For an initial-state
    $|i\rangle$ and final-state $|f\rangle$
    the scattering amplitude can be written as
    \begin{equation}\label{eqn:scattering}
        \langle f | S | i \rangle \, ,
    \end{equation}
    where the scattering matrix, or $S$-matrix
    can be decomposed into an identity matrix
    and a transfer matrix $\mathcal{T}$,
    \begin{equation}\label{eqn:S_matrix}
        S = \mathbbm{1} + \mathrm{i}\mathcal{T} \, .
    \end{equation}
    The $S$-matrix encodes all the information
    about how the initial-state will evolve
    into the final-state over time. By writing
    the $S$-matrix in this way, all the interactions
    are separated into the transfer matrix $\mathcal{T}$,
    as the identity matrix describes the free theory.
    By imposing a momentum conserving constraint
    on the $S$-matrix, the matrix element,
    $\mathcal{M}$, is defined in the following expression
    \begin{equation}\label{eqn:matrix_element}
        \langle f | \mathcal{T} | i \rangle = (2\pi)^{4} \delta^{4}\left(p_{a} + p_{b} - \sum_{f=1}^{n} p_{f}^{\mu}\right) \mathcal{M} \, ,
    \end{equation}
    where we take $p_{a}$ and $p_{b}$ to be
    the momenta of the two colliding
    particles in the initial-state, and $p_{f}$ to be the
    momenta of the $n$-particle final states.
    The transition probability of this scattering process
    is the modulus squared of the scattering amplitude
    \begin{equation}\label{eqn:S_prob}
        P = |\langle f | S | i \rangle|^{2} \propto |\langle f | \mathcal{M} | i \rangle 
    \end{equation}
    where $|\langle f | \mathcal{M} | i \rangle|^{2}  \equiv |\mathcal{M}|^{2}$
    is the matrix element squared.

    Matrix elements can be calculated from
    first principles in QFT using Feynman diagrams.

    Since detectors at collider experiments
    have limited resolution, single scattering
    events are not measured. Instead we consider
    the cross-section, $\sigma$, which is a property
    of the particles being scattered, and is independent
    of the way the experiment is carried out.
    The cross-section directly relates the
    matrix elements which we calculate in our QFT,
    to measurable observables at experiments.
    In practice, it is more useful to consider the
    differential cross-section, $\mathrm{d}\sigma$,
    where the cross-section can be differential
    in quantities such as energies and angles.
    This is because the final-state particles
    will be produced with a range of, for example,
    energies and momenta so the cross-section would
    be vanishing for an exact momentum configuration.

    The differential cross-section can be written as
    \begin{equation}\label{eqn:dsigma}
        \mathrm{d}\sigma = \dfrac{1}{\mathcal{F}}|\mathcal{M}_{2 \rightarrow n}|^{2} \mathrm{d}\Phi_{n} \, .
    \end{equation}
    $\mathcal{F}$, known as the flux factor
    is
    \begin{equation}\label{eqn:flux}
        \mathcal{F} = 4\sqrt{(p_{a}p_{b})^{2} - m_{a}^{2}m_{b}^{2}} \xrightarrow[\mathrm{limit}]{\mathrm{massless}} 2(p_{a} + p_{b})^{2} = 2s_{ab} \, .
    \end{equation}

    $\mathrm{d}\Phi_{n}$ is the Lorentz-invariant
    phase-space which contains all the possible
    configurations of the $n$-particle final state.
    Absorbing the momentum conserving $\delta$-function
    from (\ref{eqn:matrix_element}),
    we can write it as
    % \begin{equation}\label{eqn:dlips}
    %     \mathrm{d}\Phi_{n} = (2\pi)^{d}\delta^{(d)}\left(p_{a} + p_{b} - \sum_{f} p_{f}^{\mu}\right) \prod_{f=1}^{n} \dfrac{\mathrm{d}^{d}p_{f}}{(2\pi)^{d-1}}\delta(p_{f}^{2} - m_{f}^{2})\theta(E_{f}) \, ,
    % \end{equation}
    \begin{equation}\label{eqn:dlips_4d}
        \mathrm{d}\Phi_{n} = (2\pi)^{4}\delta^{4}\left(p_{a} + p_{b} - \sum_{f=1}^{n} p_{f}^{\mu}\right) \prod_{f=1}^{n} \dfrac{\mathrm{d}^{4}p_{f}}{(2\pi)^{3}}\delta(p_{f}^{2} - m_{f}^{2})\theta(E_{f}) \, ,
    \end{equation}
    where the second $\delta$-function restricts
    the final state particles to be on mass-shell
    (on-shell). The step function selects only
    the positive energy solution for this constraint.

>>>>

\section{Introduction to quantum chromodynamics}\label{sec:qcd}
    \subsection{The QCD Lagrangian}\label{sec:L_qcd}
    Quantum chronodynamics (QCD) is the sector of the SM that
    describes the strong interaction. 
    QCD is a non-Abelian gauge theory with gauge group
    $\mathrm{SU}(3)_{c}$ where the charge is named colour.
    The gauge and fermion part of the QCD Lagrangian is
    \begin{equation}\label{eqn:L_QCD}
        \mathcal{L}_{\mathrm{QCD}} = -\dfrac{1}{4}F^{a}_{\mu\nu}F^{a, \, \mu\nu}
        + \sum_{f} \bar{\psi}_{i}^{f}(\mathrm{i}\slashed{D}_{ij} - \delta_{ij}m_{f})\psi_{j}^{f} \, ,
    \end{equation}
    where twice repeated indices are summed over.
    The fields $\psi_{i}^{f}$ ($\bar{\psi}_{i}^{f}$) are the fermions (antifermions),
    representing quarks (antiquarks) with flavours $f$: 
    up $u$, down $d$, strange $s$, charm $c$, top $t$, and bottom $b$,
    with masses $m_{f}$. They transform under the fundamental
    representation with indices $i, j \in \{1, 2, 3\}$, named colour indices.
    The gauge fields $A^{a}_{\mu}$, corresponding to gluons, 
    appear in the quark covariant derivative
    \begin{equation}\label{eqn:covariant_deriv}
        (D_{\mu})_{ij} = \delta_{ij}\partial_{\mu} - ig_{\mathrm{s}}T^{a}_{ij}A^{a}_{\mu} \, ,
    \end{equation}
    and the gauge field strength tensor
    \begin{equation}\label{eqn:F_munu}
        F^{a}_{\mu\nu} = \partial_{\mu}A^{a}_{\nu} - \partial_{\nu}A^{a}_{\mu} + g_{\mathrm{s}} f^{abc}A^{b}_{\mu}A^{c}_{\nu} \, .
    \end{equation}
    Gauge fields transform under the adjoint representation, 
    which is an $N_{c}^{2}-1$ dimensional representation, meaning
    the adjoint indices $a,b,c \in \{1,...,8\}$. $T^{a}_{ij}$ are the
    group generators in a fundamental representation. In SU(3) it is
    common to write the group generators as
    \begin{equation}\label{eqn:group_generators}
        T^{a}_{ij} = \dfrac{1}{2}\lambda^{a}_{ij}
    \end{equation}
    where $\lambda^{a}_{ij}$ are the Gell-Mann matrices \cite{Gell-Mann:1962yej}.
    The generators of the group have to obey the Lie algebra
    \begin{equation}\label{eqn:lie_algebra}
        [T^{a}, T^{b}] = \mathrm{i}f^{abc}T^{c} \, ,
    \end{equation}
    where $f^{abc}$ are the structure constants of SU(3).
    The gauge coupling of the group $g_{\mathrm{s}}$
    is a dimensionless free parameter of the theory.
    It is common to use the strong coupling constant
    instead, 
    \begin{equation}\label{eqn:alpha_s}
        \alpha_{\mathrm{s}} = \dfrac{g_{\mathrm{s}}^{2}}{4\pi} \, .
    \end{equation}
    The strong coupling constant, is in fact not constant, but
    is dependent on the energy scale of the process. This
    is due to the process of renormalisation which will be discussed
    in section ...
    The implications of this is that at collider
    experiments where collisions are extremely high energy,
    the strong coupling constant becomes small, a
    propery known as asymptotic freedom. A consequence
    of this is that predictions made in QCD can be expressed
    in the form of perturbative expansions in $\alpha_{\mathrm{s}}$.

@@

    \subsection{QCD Feynman rules}\label{sec:qcd_feynman}
    Now that we have written down the QCD Lagrangian,
    we can examine the terms to write down how the
    gauge bosons interact with the fermions. A convenient
    way to do this is with Feynman rules, which can be
    pieced together to form Feynman diagrams. We will see
    that Feynman diagrams are a simple, but powerful, tool
    to systematically build all the possible ways a process
    can occur. This will be discussed further in the
    scattering amplitude section.

@@

    {\color{red} Need to write down the Feynman
    rules and draw diagrams. Best to put it in a
    figure that we can refer back to.}

    Note that we have only written down the vertices,
    propagators, and external lines for the physical gluons and quarks,
    ghosts also have associated Feynman rules but we do
    not collect those here. A full list of Feynman rules
    in the SM can be found at Ref. \cite{Romao:2012pq}.

    From these Feynman rules, we identify that each
    quark-antiquark-gluon and triple gluon vertices
    are associated with one power of $\alpha_{\mathrm{s}}$,
    and the four gluon vertex is associated with
    two powers of $\alpha_{\mathrm{s}}$. This relationship
    between vertices and $\alpha_{\mathrm{s}}$ allows
    us to systematically build up Feynman diagrams
    that correspond to a fixed order in $\alpha_{\mathrm{s}}$.
    This point will be elaborated on in Section
    \ref{sec:xs} where we discuss scattering amplitudes
    and how they are related to Feynman diagrams.

\section{Factorisation theorem}\label{sec:factorisation}
    At the LHC, collisions occur between two protons,
    which have constituent quarks and gluons, collectively
    named partons.
    The application of QCD to describe phenomenon
    in these proton-proton collisions rests on the use of
    the factorisation theorem which enables the
    separation of soft and hard energy scales.
    Within the proton, there are composite quarks
    and gluons that are constantly absorbed and
    emitted on a timescale inversely proportional
    to the mass of the proton (or hadron in general).
    The timescale of this fluctuation is much longer
    than the timescale of the interaction between
    a parton and a highly energetic probing parton
    from another proton. This is the scenario
    that occurs at collisions at the LHC.
    During the collision, or the hard scattering,
    the probe is able to interact with a
    parton that is effectively frozen.
    The probing parton knows nothing of the
    proton being probed, except for the fact
    that it collided with a parton carrying
    a fraction of the proton momentum. The
    distribution of partons within a proton
    is process independent and a fundamental
    property of the proton, meaning it can be
    separated from the actual scattering between
    the partons.

    This heuristic argument motivates the form of
    the factorisation equation, where the production of
    particles from a hadronic collision, the hadronic
    cross-section, can be written as
    \begin{equation}\label{eqn:hadronic_cs}
        \sigma_{AB \rightarrow n} = \sum_{a, b}\int_{0}^{1} \mathrm{d}x_{a}\mathrm{d}x_{b} \; f_{a/A}(x_{a},\mu_{\mathrm{F}})f_{b/B}(x_{b}, \mu_{\mathrm{F}}) \; \hat{\sigma}_{ab \rightarrow n}(Q, \mu_{\mathrm{F}}, \mu_{\mathrm{R}}) + \mathcal{O}\left(\dfrac{\Lambda_{\mathrm{QCD}}}{Q}\right)\, ,
    \end{equation}
    where $a$ is the parton from hadron $A$,
    and $b$ is the parton from hadron $B$.
    The sum over initial-state partons $a$ and $b$
    indicates the sum over flavours, which depends
    on the hadron composition in general.
    This factorisation is not exact as indicated by the
    correction term inversely proportional to characteristic
    hard scale $Q$. However, for high energy collisions, where
    $Q \gg \Lambda_{\mathrm{QCD}}$, this term and higher order
    terms are suppresed.
    In fact, the factorisation equation has only been
    proven for a few specific cases (DIS etc, cite here).

    $f_{a/h}(x, \mu)$ are parton distribution functions
    (PDFs) which depend on the momenta fractions $x$ parton $a$
    has with respect to its parent hadron $h$, at a scale $\mu$,
    usually taken to be the factorisation scale $\mu_{\mathrm{F}}$,
    the interface between soft and hard physics.
    The interpretation of PDFs at leading order are
    as probability distributions. $f_{a/h}(x, \mu)$ is
    the probability of finding parton $a$ within $h$
    carrying a momentum fraction $x$ at the energy scale $\mu$.
    PDFs are non-perturbative objects that encapsulate
    the soft effects of the scattering which occur below energy
    $\mu_{\mathrm{F}}$. They are non-perturbative because
    they are determined by fitting to experimental data,
    instead of calculated in perturbation theory.
    Multiple groups release PDF fits (cite groups here).
    In practice, these PDF fits
    are accessed through the LHAPDF interface (cite here).
    The evolution of PDFs between factorisation scales is possible
    through the use of the Dokshitzer-Gribov-Lipatov-Altarelli-Parisi (DGLAP)
    equations (cite here).

    The momenta fractions, $x$, also link the
    squared centre-of-mass energies of the hadronic collision,
    denoted as $s$, to the partonic equivalent as
    \begin{equation}\label{eqn:E_cm_partonic}
        \hat{s} = x_{a}x_{b}s \, .
    \end{equation}

    $\hat{\sigma}_{ab \rightarrow n}(Q, \mu_{\mathrm{F}}, \mu_{\mathrm{R}})$
    is the partonic cross-section which descibes the interaction of
    partons $a$ and $b$ scattering into $n$ particles,
    where the scattering occurs at energy scale $Q$,
    which is often taken to the $\hat{s}$.
    Notice that the partonic cross-section depends on
    both the factorisation scale, $\mu_{\mathrm{F}}$ and
    the renormalisation scale, $\mu_{\mathrm{R}}$
    (see \ref{chapter:qcd}).
    {\color{red} Need to mention how these scales
    are unphysical but there is a residual dependence
    on them at fixed-order calculations. Scale
    variations are a conventional way to determine
    the effects of higher order terms.
    Renormalisation will be discussed in Chapter 2
    when divergences are introduced.}

    The discussion so far has been focused hadron-hadron
    initiated scattering, however, a similar argument
    can be made about electron-positron annihilation,
    where the electron-positron annihilates to form a
    photon ($Z$ boson) which fluctuates into a quark-antiquark
    pair. For a sufficiently high energy collision,
    these interactions can be calculated in perturbation
    theory.
    Since electrons and positrons are not composite
    particles, there is no requirement for PDFs,
    therefore (\ref{eqn:hadronic_cs}) simplifies
    to just the partonic cross-section, which
    will be computed in perturbation theory.

    The details of how $\hat{\sigma}_{ab \rightarrow n}$
    is computed is discussed in the next section.

\section{Scattering amplitudes and cross-sections}\label{sec:xs}
    As mentioned in the previous section, at
    collider experiments we typically collide
    two beams, consisting of bundles of energetic
    particles, and analyse the resulting products
    of these collisions. The predictions we make
    from our theoretical model are the partonic
    cross-sections. To arrive at an expression
    for the partonic cross-section, we need to
    first consider the hard scattering of particles.
    To model the hard collisions in
    a collider experiment, we look at the specific
    case of two particles colliding into $n$ particles.
    To mathematically describe these
    scattering processes, scattering amplitudes are used.
    For an initial-state
    $|i\rangle$ and final-state $|f\rangle$
    the scattering amplitude can be written as
    \begin{equation}\label{eqn:scattering}
        \langle f | S | i \rangle \, ,
    \end{equation}
    where the scattering matrix, or $S$-matrix
    can be decomposed into an identity matrix
    and a transfer matrix $\mathcal{T}$,
    \begin{equation}\label{eqn:S_matrix}
        S = \mathbbm{1} + \mathrm{i}\mathcal{T} \, .
    \end{equation}
    The $S$-matrix encodes all the information
    about how the initial-state will evolve
    into the final-state over time. By writing
    the $S$-matrix in this way, all the interactions
    are separated into the transfer matrix $\mathcal{T}$,
    as the identity matrix describes the free theory.
    By imposing a momentum conserving constraint
    on the $S$-matrix, the matrix element,
    $\mathcal{M}$, is defined in the following expression
    \begin{equation}\label{eqn:matrix_element}
        \langle f | \mathcal{T} | i \rangle = (2\pi)^{4} \delta^{4}\left(p_{a} + p_{b} - \sum_{f=1}^{n} p_{f}^{\mu}\right) \mathcal{M} \, ,
    \end{equation}
    where we take $p_{a}$ and $p_{b}$ to be
    the momenta of the two colliding
    particles in the initial-state, and $p_{f}$ to be the
    momenta of the $n$-particle final states.
    The transition probability of this scattering process
    is the modulus squared of the scattering amplitude
    \begin{equation}\label{eqn:S_prob}
        P = |\langle f | S | i \rangle|^{2} \propto | \langle f | \mathcal{M} | i \rangle |^{2}
    \end{equation}
    where $|\langle f | \mathcal{M} | i \rangle|^{2}  \equiv |\mathcal{M}|^{2}$
    is the matrix element squared \footnote{from henceforth we refer to the matrix
    element squared as matrix element}.

    Since detectors at collider experiments
    have limited resolution, single scattering
    events are not measured. Instead we consider
    the cross-section, $\sigma$, as introduced in
    (\ref{eqn:hadronic_cs}). The cross-section is a property
    of the particles being scattered, and is independent
    of the way the experiment is carried out.
    Since the hadronic cross-section factorises
    the soft energy effects into the non-perturbative PDFs,
    which are determined once and for all for all
    processes, the object of interest now becomes the
    partonic cross-section, $\hat{\sigma}$, which
    encodes the scattering information of the specific
    process considered.

    Given that cross-sections are measurable
    quantities, the partonic cross-section directly
    relates the matrix elements which we calculate in our QFT,
    to measurable observables at experiments.
    In practice, it is more useful to consider the
    differential cross-section, $\mathrm{d}\hat{\sigma}$,
    where the cross-section can be differential
    in quantities such as energies and angles.
    This is because the final-state particles
    will be produced with a range of, for example,
    energies and momenta so the cross-section would
    be vanishing for an exact momentum configuration.

    The differential partonic cross-section can be written as
    \begin{equation}\label{eqn:dsigma}
        \mathrm{d}\hat{\sigma} = \dfrac{1}{\mathcal{F}}|\mathcal{M}_{2 \rightarrow n}|^{2} \mathrm{d}\Phi_{n} \, .
    \end{equation}
    $\mathcal{F}$, the flux factor is
    \begin{equation}\label{eqn:flux}
        \mathcal{F} = 4\sqrt{(p_{a}p_{b})^{2} - m_{a}^{2}m_{b}^{2}} \, ,
    \end{equation}
    which in the massless limit reduces to
    \begin{equation}\label{eqn:massless_flux}
        \mathcal{F} = 2(p_{a} + p_{b})^{2} = 2\hat{s} \, .
    \end{equation}
    It is common to use the massless limit because
    the centre-of-mass energy of a collider
    experiment is much greater than the mass
    of the colliding particles.

    $\mathrm{d}\Phi_{n}$ is the Lorentz-invariant
    phase-space which contains all the possible
    configurations of the $n$-particle final state.
    Absorbing the momentum conserving $\delta$-function
    from (\ref{eqn:matrix_element}),
    we can write it as
    % \begin{equation}\label{eqn:dlips}
    %     \mathrm{d}\Phi_{n} = (2\pi)^{d}\delta^{(d)}\left(p_{a} + p_{b} - \sum_{f} p_{f}^{\mu}\right) \prod_{f=1}^{n} \dfrac{\mathrm{d}^{d}p_{f}}{(2\pi)^{d-1}}\delta(p_{f}^{2} - m_{f}^{2})\theta(E_{f}) \, ,
    % \end{equation}
    \begin{equation}\label{eqn:dlips_4d}
        \mathrm{d}\Phi_{n} = (2\pi)^{4}\delta^{4}\left(p_{a} + p_{b} - \sum_{f=1}^{n} p_{f}^{\mu}\right) \prod_{f=1}^{n} \dfrac{\mathrm{d}^{4}p_{f}}{(2\pi)^{3}}\delta(p_{f}^{2} - m_{f}^{2})\theta(E_{f}) \, ,
    \end{equation}
    where the second $\delta$-function restricts
    the final state particles to be on mass-shell
    (on-shell). The step function selects only
    the positive energy solution for this constraint.
    This gives an intuitive picture of the partonic
    cross-section as the probability of a $2\rightarrow n$
    process occurring, summed over all the possible
    valid final-state configurations.

    From (\ref{eqn:S_prob}), it becomes clear
    that the matrix element is the object which
    relates back to the Lagrangian of the theory
    as it encodes the interactions of the particles.
    We saw in Section \ref{sec:qcd_feynman} that
    interactions could be codified into Feynman
    rules which are pieced together to construct
    Feynman diagrams. In this picture, matrix elements
    are exactly the sum over all Feynman diagrams
    \footnote{technically the interference of
    Feynman diagrams as we are working on the level
    of $|\mathcal{M}|^{2}$}
    for a particular process. However, there are
    an infinite number of Feynman diagrams you can
    draw as there is no constraints on how the
    vertices and propagators can be connected.
    Fortunately because $\alpha_{\mathrm{s}}$
    is small in high energy collisions, we can expand
    the matrix element as a perturbative series in
    $\alpha_{\mathrm{s}}$ to write
    \begin{equation}\label{eqn:matrix_element}
        |\mathcal{M}|^{2} =  \alpha_{\mathrm{s}}^{m} |\mathcal{M}|^{2}_{\mathrm{LO}} + \alpha_{\mathrm{s}}^{m+1} |\mathcal{M}|^{2}_{\mathrm{NLO}} + \alpha_{\mathrm{s}}^{m+2} |\mathcal{M}|^{2}_{\mathrm{NNLO}} + \mathcal{O}(\alpha_{\mathrm{s}}^{m+3}) \, ,
    \end{equation}
    where $m$ corresponds to the powers of $\alpha_{\mathrm{s}}$
    in the simplest Feynman diagrams you can draw
    for the process of interest.
    Since each term of this expansion is at a fixed-order
    in $\alpha_{\mathrm{s}}$, it is possible to systematically
    compute the diagrams which contribute at the given order of $\alpha_{\mathrm{s}}$.
    The set of diagrams that contribute at the lowest
    order in $\alpha_{\mathrm{s}}$ are $|\mathcal{M}|^{2}_{\mathrm{LO}}$,
    where LO stands for leading order. The next term
    in the expansion corresponds to the leading order
    diagrams with an additional loop, or external leg,
    which contributes an extra factor of $\alpha_{\mathrm{s}}$.
    This term is dubbed NLO for next-to-leading order
    in $\alpha_{\mathrm{s}}$, and the following term
    next-to-next-to-leading order has again an additional
    loop or leg. In general, we have terms
    N$^{k}$LO where each additional power of $\alpha_{\mathrm{s}}$
    increases complixity of the computations, however,
    because the expansion $\alpha_{\mathrm{s}}$
    is small, each additional term contribute less.
    This means that the first terms dominate the expansion,
    and it is sufficient to terminate the series after
    a few terms to reach an acceptable level of accuracy.
    A further discussion of this point will be given
    in Chapter~\ref{chapter:qcd} where we break down
    the computation of each term in (\ref{eqn:matrix_element}).

    In summary, the partonic cross-section of a collision
    reduces to the computation of the matrix elements
    up to a fixed-order in the coupling expansion,
    which is then integrated over the valid phase-space of
    final-state particle configurations. For hadronic
    collisions the partonic cross-section also needs to
    be convoluted with the PDFs to obtain the hadronic
    cross-sections.

    With the introduction of matrix elements and
    cross-sections complete, we will move the discussion onto
    Monte Carlo event generators (MCEG). These event
    generators are multi-purpose tools that
    incorporate many aspects of the theory discussed
    so far (and to be discussed) for practical use.
    Most importantly, they are
    used to compare theoretical predictions to
    experimental measurements.
